### 2013-07-01 Tex/chapters/model.tex
Korrekturen

    The Gabriel graph \cite{Gabriel1969} is a subgraph of the
    Delaunay triangulation. Two nodes \(i\) and \(j\) with distance
    \(d_{ij}\) are connected with an edge, if a circle with it's
    center on the middlepoint between \(i\) and \(j\) and radius
    \(r = \frac d 2\) contains no other nodes. This area will be
    called \emph{lune} in the following. See also Figure
    \ref{fig:lunes}\subref{sfig:lunes:def}.\\
    The Relative Neighborhood graph \cite{Toussaint1980} is a
    subgraph of the Gabriel graph. Two nodes \(i\) and \(j\) with
    distance \(d_{ij}\) are connected, if no other node is in the
    \emph{lune}. The lune is defined as the intersection of two
    circles with radius \(r = d\) and centers on \(i\) and \(j\).
    See also Figure \ref{fig:lunes}\subref{sfig:lunes:def}.
    \begin{figure}[htbp]
        \centering
        \subfigure[Definition of the lunes][]{
            \label{sfig:lunes:def}
            \input{images/GG_RNG_def}
        }
        \subfigure[Relative Neighborhood graph example][]{
            \label{sfig:lunes:rng}
            \includegraphics[width=0.3\textwidth]{images/RNG/L12S03.pdf}
        }
        \subfigure[Gabriel graph example][]{
            \label{sfig:lunes:gg}
            \includegraphics[width=0.3\textwidth]{images/GG/L12S03.pdf}
        }
        \caption[Gabriel - and Relative Neighborhood Graph]
        {
            \subref{sfig:lunes:def} Lunes of Relative Neighborhood
                Graph (hatched region) and
                Gabriel Graph (cross hatched region)
            \subref{sfig:lunes:rng} Example of a Relative
                Neighborhood Graph on periodic boundary conditions.
                Periodic nodes are dashed.
            \subref{sfig:lunes:gg} Example of a Gabriel Graph on
                periodic boundary conditions. Periodic nodes are
                dashed.
        }
        \label{fig:lunes}
    \end{figure}\\
    To construct these graphs the simple way is to test for each
    pair of nodes one has to check for every other node if it lies in
    the lune of the pair. That is of complexity \(O (N^3)\), because
    there are \(N(N-1)\) pairs and for each \(N-2\) nodes to test. So
    the product is of order \(O(N^3)\)\\
    To reduce the complexity one can first create a Delaunay
    Triangulation in complexity \(O (N \log N)\)
    \cite{Leach1992} and test the criterion for each edge, because
    the Delaunay triangulation is a supergraph of both. But the
    implementation of a Delaunay triangulation algorithm is not trivial
    and the generation of the graphs is not time critical in the scope
    of this bachelor thesis.\\
    So a tradeoff is to use basicly the simple method but only test
    the criterion for nodes which are near to the lune and abort if
    one node inside the lune is found. To determine which nodes are
    near the lune one can subdivide the area in \emph{cells} and save
    for each cell a list with nodes lying inside it like presented in
    \cite{RNGCell}.
    Now it is just neccessary to test the nodes in the cells which
    resemble a rectangular bounding box of the lune. Most pairs will be
    far away from each other and the cells in the middle of the bounding
    box are completely inside the lune so that only one node has to be
    tested to discard an edge between them. Connected nodes are near to
    each other so that only very few cells have to be tested.\\
    Indeed this method reduced the time needed to construct a Relative
    Neighborhood graph with \(N=32^2\) and \(N=64^2\) by a factor of
    over \(15\) respectivly \(40\). Though the complexity is still of
    order \(O(N^2)\) in the best case, because for every pair at least
    one check has to be performed.

>>>>

    The Gabriel graph \cite{Gabriel1969} is a subgraph of the
    Delaunay triangulation. Two nodes \(i\) and \(j\) with distance
    \(d_{ij}\) are connected with an edge, if a circle with it's
    center on the middlepoint between \(i\) and \(j\) and radius
    \(r = \frac d 2\) contains no other nodes. This area will be
    called \emph{lune} in the following. See also Figure
    \ref{fig:lunes}\subref{sfig:lunes:def}.\\
    The Relative Neighborhood graph \cite{Toussaint1980} is a
    subgraph of the Gabriel graph. Two nodes \(i\) and \(j\) with
    distance \(d_{ij}\) are connected, if no other node is in the
    \emph{lune}. The lune is defined as the intersection of two
    circles with radius \(r = d\) and centers on \(i\) and \(j\).
    See also Figure \ref{fig:lunes}\subref{sfig:lunes:def}.
    \begin{figure}[htbp]
        \centering
        \subfigure[Definition of the lunes][]{
            \label{sfig:lunes:def}
            \input{images/GG_RNG_def}
        }
        \subfigure[Relative Neighborhood graph example][]{
            \label{sfig:lunes:rng}
            \includegraphics[width=0.3\textwidth]{images/RNG/L12S03.pdf}
        }
        \subfigure[Gabriel graph example][]{
            \label{sfig:lunes:gg}
            \includegraphics[width=0.3\textwidth]{images/GG/L12S03.pdf}
        }
        \caption[Gabriel - and Relative Neighborhood Graph]
        {
            \subref{sfig:lunes:def} Lunes of Relative Neighborhood
                Graph (hatched region) and
                Gabriel Graph (cross hatched region)
            \subref{sfig:lunes:rng} Example of a Relative
                Neighborhood Graph on periodic boundary conditions.
                Periodic nodes are dashed.
            \subref{sfig:lunes:gg} Example of a Gabriel Graph on
                periodic boundary conditions. Periodic nodes are
                dashed.
        }
        \label{fig:lunes}
    \end{figure}\\
    To construct these graphs the simple way is to test for each
    pair of nodes if any other node lies in
    the lune of the pair. That is of complexity \(O (N^3)\), because
    there are \(N(N-1)\) pairs and for each \(N-2\) nodes to test. So
    the product is of order \(O(N^3)\)\\
    To reduce the complexity one can first create a Delaunay
    Triangulation in complexity \(O (N \log N)\)
    \cite{RNGCell} and test the criterion for each edge, because
    the Delaunay triangulation is a supergraph of both. But the
    implementation of a Delaunay triangulation algorithm is not trivial
    and the generation of the graphs is not time critical in the scope
    of this bachelor thesis.\\
    So a tradeoff is to use basicly the simple method but only test
    the criterion for nodes which are near to the lune and abort if
    one node inside the lune is found. To determine which nodes are
    near the lune one can subdivide the area in \emph{cells} and save
    for each cell a list with nodes lying inside it like presented in
    \cite{RNGCell}.
    Now it is just neccessary to test the nodes in the cells which
    resemble a rectangular bounding box of the lune. Most pairs will be
    far away from each other and the cells in the middle of the bounding
    box are completely inside the lune so that only one node has to be
    tested to discard an edge between them. Connected nodes are near to
    each other so that only very few cells have to be tested.\\
    Indeed this method reduced the time needed to construct a Relative
    Neighborhood graph with \(N=32^2\) and \(N=64^2\) by a factor of
    over \(15\) respectivly \(40\). Though the complexity is still of
    order \(O(N^2)\) in the best case, because for every pair at least
    one check has to be performed.

### 2013-07-01 Tex/chapters/monte_carlo.tex
Korrekturen

    \subsubsection{Wolff}
        Close to the critical temperature \(T_c\) the Metropolis
        gets slower. This is called \emph{critical slowing} down and the
        cause is beyond the scope of this thesis.\\
        Using a cluster algorithm like the Wolff
        algorithm \cite{Wolff1989} speeds things up.
        For an Ising model the Wolff algorithm builds a cluster of sites
        starting with a random site and adding neighboring sites of the
        same spin with probability \(P_{\mathrm{add}}\) from eq. \eqref{eq:wolffAdd}
        \begin{equation}
            P_{\mathrm{add}} = 1-\exp\brac{-2\beta J}
            \label{eq:wolffAdd}
        \end{equation}
        Where \(J\) is the coupling constant (explained in section
        \ref{ssec:isingmodel}). The neighboring sites of the added sites
        are also considered and so forth. When there are no more sites
        to add, the spin of every site in the cluster is flipped
        \cite[S. ??]{NewmanBarkema1999} \cite[S. 151f]{Katzgraber2011}.
        This leads fast to new uncorrelated states at the critical
        temperature because big clusters are flipped. But there are not
        much advantages at high or low temperatures. At low temperatures
        the cluster will consist of almost all sites such that all but
        very few spins will be flipped. At high temperatures the cluster
        will only contain very few sites.
        Both situaitions have no advantage against the Metropolis algorithm.\\
        So one would activate this algorithm near the critical temperature
        but use a simple Metropolis algorithm at high and low temperatures.

@@

\subsection{Equilibration- and Autocorrelation Time}
\label{ssec:eqtime}
    To generate an equilibrium state one starts with an arbitrary state
    and waits until it is equilibrated. The count of sweeps till
    equilibrium is called \emph{equilibration time} \(t_{eq}\).
    All measurement should start after this time.\\
    To determine when the system is in equilibrium, one can watch the
    development of some observables and take the point at which there
    are no big changes anymore as equilibrium. In fig.
    \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
    the equilibrium is reached after approximately \(N=50\) sweeps for
    either an initial condition of all spins up and all spins random. It
    does not harm to double that value to be save.
    \begin{figure}[htbp]
        \centering
        \subfigure[Example of an equilibrating Ising system][]{
                \label{sfig:equiandauto:equiE}
                \includegraphics[width=0.47\textwidth]{plots/equiE}
        }
        \subfigure[Example of the autocorrelation of an Ising system][]{
                \label{sfig:equiandauto:autoM}
                \includegraphics[width=0.47\textwidth]{plots/autoM}
        }
        \caption[Examples for equilibration and autocorrelation]
        {
            \subref{sfig:equiandauto:equiE} Example of a Ising system
                \(L=64\) reaching equilibrium at \(T=2.36\) and
            \subref{sfig:equiandauto:autoM} the autocorrelation of an
                Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                sweeps -- otherwise the decline is too steep to show)
                on half logarithmic axis.
                The straight line is a exponential fit \(\exp(-t/\tau)\)
                with \(\tau = 342(1)\).
        }
        \label{fig:equiandauto}
    \end{figure}\\
    Because every state is generated from the state before, measurements
    of subsequent states are correlated. To determine when two states
    are independet, one calculates the normalized autocorrelation function
    \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
    \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
    logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
    To get the autocorrelation time one can either fit a exponential
    function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
    or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
    \(\tau\) is an estimate after which time two samples are not
    correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
    To make sure that the error is not underestimated one should wait
    \(2\tau\) sweeps between two measurements.
    The autocorrelation time is of course dependent on the temperature.
    At high temperatures the fluctuations are strong and subsequent
    states are more dissimilar and therefore less correlated than at low
    temperatures, where less spins are flipping. But the longest
    autocorrelation times are encountered at the critical temperature.
    This effect is called \emph{critical slowing down} and is
    characterized by the \emph{dynamical critical exponent} \(z\)
    \cite{SwendsenWang1987}. The dependence of the autocorrelation time
    \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).

>>>>

    \subsubsection{Wolff}
    \label{sssec:wolff}
        Close to the critical temperature \(T_c\) the Metropolis
        gets slower. This is called \emph{critical slowing} down and the
        cause is beyond the scope of this thesis.\\
        Using a cluster algorithm like the Wolff
        algorithm \cite{Wolff1989} speeds things up.
        For an Ising model the Wolff algorithm builds a cluster of sites
        starting with a random site and adding neighboring sites of the
        same spin with probability \(P_{\mathrm{add}}\) from eq. \eqref{eq:wolffAdd}
        \begin{equation}
            P_{\mathrm{add}} = 1-\exp\brac{-2\beta J}
            \label{eq:wolffAdd}
        \end{equation}
        Where \(J\) is the coupling constant (explained in section
        \ref{ssec:isingmodel}). The neighboring sites of the added sites
        are also considered and so forth. When there are no more sites
        to add, the spin of every site in the cluster is flipped
        \cite[S. ??]{NewmanBarkema1999} \cite[S. 151f]{Katzgraber2011}.
        This leads fast to new uncorrelated states at the critical
        temperature because big clusters are flipped. But there are not
        much advantages at high or low temperatures. At low temperatures
        the cluster will consist of almost all sites such that all but
        very few spins will be flipped. At high temperatures the cluster
        will only contain very few sites.
        Both situaitions have no advantage against the Metropolis algorithm.\\
        So one would activate this algorithm near the critical temperature
        but use a simple Metropolis algorithm at high and low temperatures.

@@

\subsection{Equilibration- and Autocorrelation Time}
\label{ssec:eqtime}
    To generate an equilibrium state one starts with an arbitrary state
    and waits until it is equilibrated. The count of sweeps till
    equilibrium is called \emph{equilibration time} \(t_{eq}\).
    All measurement should start after this time.\\
    To determine when the system is in equilibrium, one can watch the
    development of some observables and take the point at which there
    are no big changes anymore as equilibrium. In fig.
    \ref{fig:equiandauto}\subref{sfig:equiandauto:equiE}
    the equilibrium is reached after approximately \(N=50\) sweeps for
    either an initial condition of all spins up and all spins random. It
    does not harm to double that value to be save.
    \begin{figure}[htbp]
        \centering
        \subfigure[Example of an equilibrating Ising system][]{
                \label{sfig:equiandauto:equiE}
                \includegraphics[width=0.47\textwidth]{plots/equiE}
        }
        \subfigure[Example of the autocorrelation of an Ising system][]{
                \label{sfig:equiandauto:autoM}
                \includegraphics[width=0.47\textwidth]{plots/autoM}
        }
        \caption[Examples for equilibration and autocorrelation]
        {
            \subref{sfig:equiandauto:equiE} Example of a Ising system
                \(L=64\) reaching equilibrium at \(T=2.36\) and
            \subref{sfig:equiandauto:autoM} the autocorrelation of an
                Ising system \(L=64\) at \(T=2.40\) (only Metropolis
                sweeps -- otherwise the decline is too steep to show)
                on half logarithmic axis.
                The straight line is an exponential fit \(\exp(-t/\tau)\)
                with \(\tau = 342(1)\).
        }
        \label{fig:equiandauto}
    \end{figure}\\
    Because every state is generated from the state before, measurements
    of subsequent states are correlated. To determine when two states
    are independent, one calculates the normalized autocorrelation function
    \(\frac{\chi(t)}{\chi(0)}\), which should decay exponentially
    \(\chi(t) \propto \exp(t/\tau)\). This is visible in the half
    logarithmic plot \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}.
    To get the autocorrelation time one can either fit a exponential
    function \(\exp(-t/\tau)\) like in fig \ref{fig:equiandauto}\subref{sfig:equiandauto:autoM}
    or integrate \(\tau = \int \frac{\chi(t)}{\chi(0)} \de t\).
    \(\tau\) is an estimate after which time two samples are not
    correlated anymore. \cite[S. ??]{NewmanBarkema1999} \cite[S. 150f]{Katzgraber2011}.
    To make sure that the error is not underestimated one should wait
    \(2\tau\) sweeps between two measurements.
    The autocorrelation time is of course dependent on the temperature.
    For example for the standard Metropolis algorithm the fluctuations
    are strong at high temperatures and subsequent
    states are more dissimilar and therefore less correlated than at low
    temperatures, where less spins are flipping. But the longest
    autocorrelation times are encountered at the critical temperature.
    This effect is called \emph{critical slowing down} and is
    characterized by the \emph{dynamical critical exponent} \(z\)
    \cite{SwendsenWang1987}. The dependence of the autocorrelation time
    \(\tau\) on the system size \(L\) is then given by \(\tau \propto L^z\).
    As mentioned in section \ref{sssec:wolff}, the Wolff cluster algorithm
    decreases \(z\) dramatically. This causes the course of the autocorrelation
    time in dependence on temperature to change significantly as shown in
    section \ref{sec:results}.
